---
title: "Differentially expressed protein analysis of SWATH data using R"
output: html_notebook
---

This worflow shows how to process data from SWATH-MS protein quantification, in order to study differential expression of the proteins between the different conditions.

The data in this example come from : Bjelosevic S, Pascovici D, Ping H, et.al. Quantitative age-specific variability of plasma proteins in healthy neonates, children and adults. Molecular & Cellular Proteomics (March 23 2017)

#1. loading and visualizing the data 
```{r}
data<-read.table("SWATH.example.data.csv", sep=",",header=T, row.names = 1)
data
boxplot(data, ylab="Intensity", xlab="samples") 
hist(as.numeric(unlist(data)), main = "Histogram of Intensity distribution",  xlab = "Intensity")
```

#3. Normalization

##3.1 design matrix

In order to perform the normalization with the limma package we need to define the design matrix generated with model.matrix(), which identify which sample belong to which condition.  :  


```{r}
#for this specific data :
exp.design <- data.frame(samples = colnames(data), condition = 1)
exp.design$condition[1:10] = "control"
exp.design$condition[11:20] = "lessone"
exp.design$condition[21:30] = "onetofive"
exp.design$condition[31:40] = "adult"

design <- model.matrix(~0 + exp.design$condition, data = exp.design)
colnames(design) <- sort(unique(exp.design$condition))
row.names(design) <- exp.design$samples
as.data.frame(design)
```


```{r}
contrast <- makeContrasts(adult-control, lessone-control, onetofive-control, adult+lessone+onetofive-control, levels=design)
as.data.frame(contrast)
```


##3.2 applying limma

```{r}
library(limma)
library(edgeR)

dge <- DGEList(data)
dge <- calcNormFactors(dge)
y <- voom(dge, design)
norm.data <- y$E
as.data.frame(norm.data)
boxplot(norm.data, ylab="Intensity", xlab="samples")
hist(norm.data, main = "Histogram of Intensity distribution",  xlab = "Intensity")
```

```{r}
contrast <- makeContrasts(adult-control, lessone-control, onetofive-control, levels=design)
as.data.frame(contrast)
```


#4. Differential expression annalysis
##4.1 fit model
```{r}
fit <- lmFit(y, design)
fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)
```


```{r}
adult.vs.c <- topTable(fit2, coef = "adult - control", n = nrow(fit2))
lessone.vs.control <- topTable(fit2, coef = "lessone - control", n = nrow(fit2))
onetofive.vs.control <- topTable(fit2, coef = "onetofive - control", n = nrow(fit2))
adult.vs.c ; lessone.vs.control ; onetofive.vs.control 
```
```{r}
adult.vs.c$protein <- row.names(adult.vs.c)
row.names(adult.vs.c) <- NULL
adult.vs.c <- adult.vs.c[, c(7, 4, 5, 1)]

lessone.vs.control$protein <- row.names(lessone.vs.control)
row.names(lessone.vs.control) <- NULL
lessone.vs.control <- lessone.vs.control[, c(7, 4, 5, 1)]

onetofive.vs.control$protein <- row.names(onetofive.vs.control)
row.names(onetofive.vs.control) <- NULL
onetofive.vs.control <- onetofive.vs.control[, c(7, 4, 5, 1)]

Results <- list(adult.vs.c, lessone.vs.control, onetofive.vs.control)
```




##4.2 Differentially expressed proteins

```{r}

thresh_fc <- 0.5
thresh_p <- 0.05
deProt <- Results
for(i in 1:length(Results)){
    fc = as.data.frame(deProt[[i]])[,4]
    p = as.data.frame(deProt[[i]])[,3]
    dt <-as.data.frame(deProt[[i]])
    deProt[[i]] <- dt[which(p<=thresh_p & abs(fc)>=thresh_fc),]
    print(as.data.frame(deProt[[i]]))
  }
```




##4.3 Volcano Plot

```{r}

library(ggplot2)
library(ggrepel)

tresh_fc = 0.5 ## Fold change threshold
tresh_p = 0.05 ## p.value treshold

for(i in 1:length(Results)){
  
  plotTitle <- substr(colnames(Results[[i]])[2], 9 ,nchar(colnames(Results[[i]])[2]))
  values <- as.data.frame(Results[[i]])
  forplot <- data.frame(x=as.numeric(values[,4]), y=-log10(values[,3]), id=as.character(values[,1]))
  tmp <- forplot[as.numeric(forplot$y)>=-log10(tresh_fc) & abs(forplot$x)>tresh_fc,]
  p <- ggplot(forplot) + geom_point(aes(x, y , color = ifelse(y>=-log10(tresh_p) & abs(x)>=tresh_fc, "not signi", "FC")),show.legend = F) +
    scale_color_manual(values = c("blue", "red")) +
    geom_text_repel(data = subset(forplot, abs(forplot$x)>=tresh_fc & forplot$y>=-log10(tresh_p)),
                    aes(x,y,label = id),
                    size = 2) +
    geom_vline(xintercept = tresh_fc ) +
    geom_vline(xintercept = -tresh_fc) + 
    geom_hline(yintercept = -log10(tresh_p)) + 
    labs(title = plotTitle,x="log2(Fold-change)", y="-log10(P.Value)") + theme_bw() 
    print(p)
}
```


```{r}
boxplot(apply(norm.data,1,sd),apply(scale(log2(data)),1,sd),main = "comparison of the standard deviation \n of the differents normalization techniques", xlab="technique", ylab="standard deviation",names=c("voom", "center and scaling"))
```


